MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments \textbf{Supplementary Information}

### 2022-07-28 main/MSstatsPTM_Main.tex
addressed all comments in main

Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of biomolecular events. Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen,Bludau2022}. Characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}. For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}. Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Chandramouli:2009}\cite{Kim2016}. This helps to distinguish between relative site occupancy changes at steady-state protein levels, typical for short time-scale signaling events, and observed relative changes of PTMs as a result of underlying gene expression or protein abundance levels.

Bottom-up liquid chromatography coupled with tandem mass spectrometry (LC-MS/MS) is a tool of choice for unbiased and large-scale identification and quantification of proteins and their PTMs~\cite{Kall:2011ub,Roepstorff}. However, LC-MS-based interrogation of the modified proteome is challenging, for a number of reasons. First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \cite{Huang:2014}. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities. Second, contrary to the often large number of identified peptides that can be used to quantify protein abundance, there are relatively few representative peptides that span a modification site, and there may be multiple modified sites on a single peptide~\cite{Mann}. Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complicating the interpretation of the results \cite{Olsen:2013}\cite{Wu:2011}. Finally, technological aspects of bottom-up MS experiments, such as presence of labeling by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.

@@

\todo{I do not think this is right. I believe there should be a summarization first, such that the data are $\hat{y}_{ij}$, the model becomes $\hat{y}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2})$ and the df are $df=I(J-1)$. The index k should drop. Could you start by introducing the summarization? What do you think re keeping or removing the hat on $\hat{y}_{ij}$?}
The basic Analysis of Variance (ANOVA) model~\cite{kutner_etal_04a}  is
\begin{eqnarray}
y^*_{ijk} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}$ and its standard error, and test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(JK-1)$ \todo{$df=I(J-1)$} degrees of freedom in balanced designs.
Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.

@@

\medskip \noindent 
The basic ANOVA can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}. Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g. with the sum on the original scale, and considering sample-wise ratios of the summaries. On the log$_2$ scale, for condition $i$ and biological replicate $j$ the approach is equivalent to calculating
\begin{eqnarray}
u_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
\label{eq:adj_estimation}
\end{eqnarray}
The approach then models these values with the basic ANOVA, which corresponds to \todo{if we keep a hat on $\hat{y}_{ij}$, we should probably keep a hat on $\hat{u}_{ij}$}
\begin{eqnarray}
u_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.

@@

\medskip \noindent The estimation of nuisance variation of the ratio-based ANOVA model above is often further expanded with Empirical Bayes moderation implemented in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}. A typical application of Limma takes as input $u_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_model}. 
The model differs from~\eqref{ttest_model} in that it specifies additional prior distributions for the model parameters.
The prior distributions are estimated from the data across all the proteins and all the PTM as described in \todo{XXX}. 
With this approach, testing the null hypothesis in~\eqref{null_hyp} amounts to combining the PTM- and protein-specific estimates of variation with a consensus estimate obtained from all the PTM and all the proteins. \todo{It's a bit vague. I am wondering if an equation would help}
As the result, in experiments with few biological replicates the standard errors are often smaller, and the degrees of freedom are often larger than without moderation~\cite{Ritchie_15a}. 
Thus the approach tends to increase the sensitivity of detecting differential abundance.

@@

\medskip \noindent Isobar-PTM was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as TMT, or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}. Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. \todo{This seems somewhat similar to the ANOVA on ratios. Could you contrast with the ANOVA? Or just provide an equation in the same notation?} Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.

@@

$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a family of R/Bioconductor packages for statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. The packages take as input log$_2$-intensities $y_{ijk}$. For each protein, the log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}. The summaries are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}. The specific model fit for the individual protein depends on the design of the experiment, labeling type and data acquisition type, as well as on the pattern of missing values in that particular protein. For example, the unmodified protein features in \figref{data-structure} are modeled as
\todo{I rewrote the model so that it uses $\mu_i$ directly rather than conditions}
\todo{Should we use a hat on top of y, to indicate that we model the summaries?}
\begin{eqnarray}
& y_{ij} = \mu_i + Run_{j} + \epsilon_{ij}, \  \text{where } Run \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2_j), \epsilon_{ij} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2) & 
\label{eq:msstats_model}
\end{eqnarray}
\todo{Is it true that we add run in label-free experiments? I thought it was only with TMT. But maybe I forgot. Could you check? If we don't include run, then give a model for label-free, and then contrast with TMT}
\todo{Is it true that we have a separate variance parameter for each run $ \sigma^2_j$? I don't think so. I think it should be a separate sigma (maybe denote it $\tau^2$), but same for all the runs}
Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
In turn, the parameters allow us to estimate the pairwise comparison $\hat{\Delta}_{Protein} = \hat{\mu}_{i} - \hat{\mu}_{i^{\prime}}$ and its standard error. Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}, which increases the sensitivity of detecting differential abundance when the number of biological replicates in each condition is small.

@@

We leverage the existing methods in $MSstats$ and $MSstatsTMT$, and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.  Each feature type is first analyzed separately using $MSstats$ or $MSstatsTMT$. In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$ \todo{check the use of hat}. The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in \eqref{msstats_model}. The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$. Similarly, the unmodified features of each protein are summarized for each run into $\hat{y}_{ij}$, and the summaries are used as input to a separate analysis by $MSstats$ or $MSstatsTMT$ producing $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$ and $\widehat{SE}(\hat{\Delta}_{protein})$.

@@

\todo{This paragraph relates more to the implementation. Could you integrate it with the implementation section?} By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile. It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs. It is applicable to label-free data acquisition strategies such as DDA and DIA and experiments with TMT labeling. The proposed approach is also applicable to experiments where the experimental designs for PTM profiling and global proteome profiling vary in properties such as number of biological replicates, data acquisition strategies and runs.

@@

\todo{I suggest that you follow the same order as Figure 3 (my suggested version), one paragraph per step. Describe each step, and refer to both the figure, the quantity in the equations and the MSstats functions. Specifically: design; data acquisition; data processing and requirements; converters; steps of statistical analysis. I would state upfront that all the steps before last are performed separately for each data type, and not repeat it in later paragraphs}

The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}. $MSstatsPTM$ is applicable to most experimental designs, label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT. $MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.

$MSstatsPTM$ requires the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e. the modification site). This will generally include the amino acid abbreviation, plus it's number in the protein sequence. For example, a modification on a 70th amino acid in the sequence serine should be marked as ``S70''.

Occasionally the outputs of data processing tools only include the peptide sequence with the modified amino acid highlighted, without indicating where in the protein sequence the modification occurred. For these cases $MSstatsPTM$ includes functionality for identifying this location \todo{what do you mean by 'manually'? The users have to manually input it?} manually, given the modified peptide sequence and a FASTA file with the entire protein sequence. The package includes converters for seamless integration \todo{integration of what?} with multiple such tools. \todo{Maybe put all the discussion of converters in one place - integrate with the first paragraph? } The converters can optionally keep peptides with both single and multiple modification sites, or filter out only those with multiple sites. If multiple sites are kept, these are grouped separately from the singly modified peptide so as not to confound the effect \todo{confound how? The effect of what? We use confounding to indicate the changes in protein abundance. This looks like a different confounding. Could you clarify?}. Additionally, peptides with oxidation can also be removed if the effect is believed to confound the modifications of interest \todo{confound how? The effect of what?}. For experimental designs that include separate runs enriched in modified peptides and global profiling runs, conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs.

The next step is PTM/protein summarization. Summarization is performed separately for the PTM and unmodified protein. When summarizing the PTM, modified peptide features that span the same modification site are summarized together. Peptides that include multiple modifications are not included in the single modification summarization, and are grouped separately. The unmodified protein summarization is performed as discussed above for $MSstats$. When summarizing the unmodified protein features, the package optionally imputes missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}. When summarizing the modified features, missing value imputation is also possible but should be performed with care. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can violate the assumptions underlying the imputation, and lead to numerically unstable  results. Next, separate statistical models are fit to both the summaries of the modified features and the to summaries of the unmodified features. The models are automatically selected to reflect the experimental design and the data acquisition. If the base model is not applicable for a particular PTM or protein, e.g. due to missing data, a simplified model is fit.

After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above. Modification sites for which lacking the corresponding global profiling information cannot be adjusted for changes in protein abundance. In this case the implementation reverts to testing the null hypothesis in~\eqref{conv_null_hyp} using the statistical methods seen in $MSstats$, applied separately to each modified peptide. In addition to the statistical functionalities, the implementation includes visualizations for quality control and assessment of the quality of model fit.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.5]{images/MSstatsPTM_design.png}
\caption{\todo{Looks great. Maybe to the left add vertical text 'Upstream experiments and computation' and  'MSstatsPTM' next to the corresponding boxes. No need for a capital L in lysate. 'Group comparison among PTM' and 'Group comparison among unmodified proteins'. I would add () next to the function names, to clarify that these are functions} The $MSstatsPTM$ workflow. The names of the R functions used for each step are highlighted in purple and the output notations are highlighted in red. The package is applicable to label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT. It takes as input lists of identified and quantified spectral features for the PTM and unmodified protein, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.  Conversion, summarization and statistical modeling are performed separately for the PTM and for the unmodified proteins. At the end, the model-based summaries are combined to adjust the changes in the PTM abundance for changes in unmodified protein abundance. Data visualizations can optionally be created at each step.}
\label{fig:msstatsptm_design}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.7\linewidth}
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{}
\label{fig:sim1_fdr}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{}
\label{fig:sim2_acc}
\end{subfigure}
\caption{\todo{In the figure titles and in the caption, explain which simulation is 'clean' and which is 'noisy'. Use the same names for the datasets as in the Datasets section throughout. ANOVA is an abbreviation and needs all capital letters - could you check throughout in the figures} Dataset 1 \& 2: Computer simulation. a) All the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates. b) The advantage of using the proposed approach was apparent when including limited observations and missing values. Looking at accuracy, the proposed method outperformed ANOVA and Limma in nearly every model.
}
\label{fig:computer_sim}
\end{figure}

>>>>

Signaling mechanisms allow cells to mount a fast and dynamic response to a multitude of biomolecular events. Signaling is facilitated by the modification of proteins at specific residues, acting as molecular on/off switches~\cite{Deribe,Cohen,Bludau2022}. Characterizing relative abundance of a modification site's occupancy repertoire across experimental conditions provides important insights~\cite{Mann}. For example, meaningful patterns of changes in post-translational modifications (PTMs) abundance can serve as biomarkers of a disease~\cite{Petushkova_2017}. Alternatively, distinguishing the quantitative changes in a PTM from the overall changes of the protein abundance helps gain insight into biological and physiological processes operating on a very short timescale~\cite{Wu:2011, Chandramouli:2009, Kim2016}. This helps to distinguish between relative site occupancy changes at steady-state protein levels, typical for short time-scale signaling events, and observed relative changes of PTMs as a result of underlying gene expression or protein abundance levels.

Bottom-up liquid chromatography coupled with tandem mass spectrometry (LC-MS/MS) is a tool of choice for unbiased and large-scale identification and quantification of proteins and their PTMs~\cite{Kall:2011ub,Roepstorff}. However, LC-MS-based interrogation of the modified proteome is challenging, for a number of reasons. First, the relatively lower abundance of modified proteoforms dictates that a global interrogation can only be achieved through large-scale enrichment protocols with modification-specific antibodies or beads \cite{Huang:2014}. Variability in the enrichment efficiency inevitably affects the reproducibility of the number of spectral features (e.g., peptide precursor ions or their fragments) and their intensities. Second, contrary to the often large number of identified peptides that can be used to quantify protein abundance, there are relatively few representative peptides that span a modification site, and there may be multiple modified sites on a single peptide~\cite{Mann}. Third, unless early signaling events are interrogated, the interpretation of the relative changes in modification occupancy are inherently confounded with changes in the overall protein abundance, complicating the interpretation of the results \cite{Wu:2011, Olsen:2013}. Finally, technological aspects of bottom-up MS experiments, such as presence of labeling by tandem mass tag (TMT), introduce additional sources of uncertainty and variation.

@@

The basic Analysis of Variance (ANOVA) model~\cite{kutner_etal_04a}  is
\begin{eqnarray}
\hat{u}^*_{ij} = \mu^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J
\label{eq:ttest_model}
\end{eqnarray}

Where $\hat{u}^*_{ij}$ is the run-level summarization of modified features. For example, on the log$_2$ scale for condition $i$ and biological replicate $j$, $\hat{u}^*_{ij}$ is

\begin{eqnarray}
\hat{u}^*_{ij} =  \log_2 \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}}
\label{eq:anova_summ}
\end{eqnarray}

The model allows us to estimate $\hat{\Delta}$ and its standard error, and test the null hypothesis in \eqref{conv_null_hyp} by comparing the model-based test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.
Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the unmodified protein.

%\medskip \noindent

@@

\medskip \noindent 
The basic ANOVA can be extended to account for the confounding of changes in PTM abundance and overall changes in protein abundance~\cite{Schwammle2015,THOMAS2020,Mertins:2013}. Typically this is done by first summarizing the feature intensities in each run, separately for modified and unmodified features, e.g. with the sum on the original scale, and considering sample-wise ratios of the summaries. On the log$_2$ scale, for condition $i$ and biological replicate $j$ the approach is equivalent to calculating
\begin{eqnarray}
\hat{u}_{ij} = \log_2 \frac{ \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} }{ \sum_{l=1}^{L} 2^{y_{ijl}} } = \log_2 \left( \sum_{k=1}^{K} 2^{y_{ijk}^{\ast}} \right) - \log_2 \left( \sum_{l=1}^{L} 2^{y_{ijl}} \right)
\label{eq:adj_estimation}
\end{eqnarray}
The approach then models these values with the basic ANOVA, which corresponds to 
\begin{eqnarray}
\hat{u}_{ij} = (\mu^{\ast}_{i}-\mu_i) + \epsilon^{\prime}_{ij}, \text{where } \epsilon^{\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime 2}),\ i=1,\ldots,I,\ j=1,\ldots,J
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}_{adj}$ and its standard error.
Based on this model, we can test the more relevant null hypothesis in~\eqref{null_hyp}, by comparing the test statistic against the Student distribution with $df=I(J-1)$ degrees of freedom in balanced designs.

@@

\medskip \noindent The estimation of nuisance variation of the ratio-based ANOVA model above is often further expanded with Empirical Bayes moderation implemented in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}. A typical application of Limma takes as input $u_{ij}$, and for each PTM fits the linear model in~\eqref{ttest_model}. 
The model differs from~\eqref{ttest_model} in that it specifies additional prior distributions for the model parameters.
The prior distributions are estimated from the data across all the proteins and all the PTM as described in \cite{Smyth:2004}. 
With this approach, testing the null hypothesis in~\eqref{null_hyp} amounts to combining the PTM- and protein-specific estimates of variation with a consensus estimate obtained from all the PTM and all the proteins.
As the result, in experiments with few biological replicates the standard errors are often smaller, and the degrees of freedom are often larger than without moderation~\cite{Ritchie_15a}. 
Thus the approach tends to increase the sensitivity of detecting differential abundance.

@@

\medskip \noindent Isobar-PTM was also proposed for experiments with LC-MS/MS quantitative strategies that employ isobaric labels such as TMT, or isobaric tag for relative and absolute quantification (iTRAQ)\cite{Breitwieser:2013}. Isobar-PTM expresses MS measurements with a linear model and performs adjustment with respect to protein abundance using the difference between log-ratio of modified peptides in two channels and log-ratio of protein level. Unfortunately, this statistical modeling framework is not applicable to either label-free workflows or experiments with complex designs.

@@

$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a family of R/Bioconductor packages for statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. The packages take as input log$_2$-intensities $y_{ijk}$. For each protein, the log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}. The summaries are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}. The specific model fit for the individual protein depends on the design of the experiment, labeling type and data acquisition type, as well as on the pattern of missing values in that particular protein. For example, the unmodified protein features in \figref{data-structure} are modeled as

\begin{eqnarray}
\hat{y}_{ij} = \mu_i + \epsilon_{ij}, \text{where } \epsilon_{ij} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)
\label{eq:msstats_model}
\end{eqnarray}

In the simplified case of \figref{data-structure} this amounts to the same model as in \eqref{ttest_model}. However, the model can be extended to more complex cases with additional sources of variation, which simpler models, such as ANOVA, cannot account for. For example, given a group comparison TMT experiment, with the addition of mixture as a source of variation, the model would extend to

\begin{eqnarray}
\hat{y}_{imj} = \mu_i + Mixture_m + \epsilon_{imj}, \text{where } Mixture_m \stackrel{iid}{\sim} \mathcal{N}(0, \sigma_M^2), \epsilon_{imj} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2)
\label{eq:msstatstmt_model}
\end{eqnarray}

Additional models for different experimental designs, including time series, can be seen in {\bf Supplementary Figure~\ref{supp-fig:statistical_inference}}.

Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
In turn, the parameters allow us to estimate the pairwise comparison $\hat{\Delta}_{Protein} = \hat{\mu}_{i} - \hat{\mu}_{i^{\prime}}$ and its standard error. Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}, which increases the sensitivity of detecting differential abundance when the number of biological replicates in each condition is small.

@@

We leverage the existing methods in $MSstats$ and $MSstatsTMT$, and extend them to remove confounding between changes in the PTM and the overall changes in protein abundance.  Each feature type is first analyzed separately using $MSstats$ or $MSstatsTMT$. In particular, the modified features are summarized into run-level summaries $\hat{y}_{ij}^{\ast}$. The estimated summaries of the modified features are then used as the input to the appropriate model in $MSstats$ or $MSstatsTMT$,  such as in \eqref{msstats_model}. The resulting model-based estimates include $\hat{\Delta}_{PTM}=\hat{\mu}^{\ast}_{i}-\hat{\mu}_{i^{\prime}}^{\ast}$, and its standard error $\widehat{SE}(\hat{\Delta}_{PTM})$. Similarly, the unmodified features of each protein are summarized for each run into $\hat{y}_{ij}$, and the summaries are used as input to a separate analysis by $MSstats$ or $MSstatsTMT$ producing $\hat{\Delta}_{protein}=\hat{\mu}_{i}-\hat{\mu}_{i^{\prime}}$ and $\widehat{SE}(\hat{\Delta}_{protein})$.

@@

@@

The proposed methods are implemented in the open source R package $MSstatsPTM$, overviewed in~\figref{msstatsptm_design}. By leveraging the implementations in $MSstats$ and $MSstatsTMT$, the proposed approach is very versatile. It is applicable to a wide variety of experimental designs, including group comparison, paired designs, time course designs and unbalanced designs. It is applicable to label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT. It can model experiments where the experimental designs for PTM profiling and global proteome profiling vary in properties such as number of biological replicates, data acquisition strategies and runs.

$MSstatsPTM$ takes as input lists of identified and quantified spectral features, produced by spectral processing tools, such as MaxQuant, Progenesis, or Spectronaut (step 3 of~\figref{msstatsptm_design}). Conversion is performed separately for the runs enriched in modified peptides and for the global profiling runs. We require the processing tool to identify the amino acid in the protein sequence where the modification occurred (i.e. the modification site). This will generally include the amino acid abbreviation, plus it's number in the protein sequence. For example, a modification on a 70th amino acid in the sequence, serine should be marked as ``S70''. Occasionally the outputs of data processing tools only include the peptide sequence with the modified amino acid highlighted, without indicating where in the protein sequence the modification occurred. For these cases $MSstatsPTM$ includes functionality for identifying the location, given the modified peptide sequence and a FASTA file with the entire protein sequence. The converters output the modified spectral features $y_{ijk}^{\ast}$ and the corresponding unmodified features $y_{ijk}$ in the format required for summarization.

The next step is PTM/protein summarization using the $dataSummarizationPTM()$ function (step 4 of~\figref{msstatsptm_design}). Summarization is performed separately for the PTM and unmodified protein. When summarizing the PTM, modified peptide features that span the same modification site are summarized together. Peptides that include multiple modifications are not included in the single modification summarization, and are grouped separately. The unmodified protein summarization is performed as discussed above for $MSstats$. When summarizing the unmodified protein features, the package optionally imputes missing values using an Accelerated Failure Time (AFT) model~\cite{Wei:1992}. When summarizing the modified features, missing value imputation is also possible but should be performed with care. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can violate the assumptions underlying the imputation, and lead to numerically unstable results. The outputs of this step are the run-level summaries for the modified, $\hat{y}_{ij}^{\ast}$, and unmodified, $\hat{y}_{ij}$, features.

Finally, separate statistical models are fit to both feature summaries using the $groupComparisonPTM()$ function (step 5 of~\figref{msstatsptm_design}). The models are automatically selected to reflect the experimental design and the data acquisition. If the base model is not applicable for a particular PTM or protein, e.g. due to missing data, a simplified model is fit. The output of the models are the estimates $\hat{\Delta}_{PTM}$ and $\hat{\Delta}_{protein}$, and their standard errors $\widehat{SE}(\hat{\Delta}_{PTM})$ and $\widehat{SE}(\hat{\Delta}_{protein})$. After modeling, the modified model is adjusted for changes in unmodified protein abundance, using the methods described above. Modification sites for which lacking the corresponding global profiling information cannot be adjusted for changes in protein abundance. In this case the implementation reverts to testing the null hypothesis in~\eqref{conv_null_hyp} using the statistical methods seen in $MSstats$, applied separately to each modified peptide. The final output is the estimate $\hat{\Delta}_{adj}$ and it's standard error $\widehat{SE}(\hat{\Delta}_{adj})$.

In addition to the above functionalities, the implementation includes visualizations for quality control, $dataProcessPlotsPTM()$, and assessment of the quality of model fit, $groupComparisonPlotsPTM()$.

@@

\begin{figure}[ht]
\centering
\includegraphics[scale=.5]{images/MSstatsPTM_design.png}
\caption{The $MSstatsPTM$ workflow. The names of the R functions used for each step are highlighted in purple and the output notations are highlighted in red. The workflow begins with the acquisition of the enriched and global proteome lysates. The package is applicable to label-free data acquisitions such as DDA, DIA, SRM and label-based data acquisitions such as TMT. It takes as input lists of identified and quantified spectral features for the PTM and unmodified protein, produced by spectral processing tools such as MaxQuant, Progenesis, or Spectronaut.  Conversion, summarization and statistical modeling are performed separately for the PTM and for the unmodified proteins. At the end, the model-based summaries are combined to adjust the changes in the PTM abundance for changes in unmodified protein abundance. Data visualizations can optionally be created after steps 4 and 5 of the workflow.}
\label{fig:msstatsptm_design}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.7\linewidth}
\includegraphics[width=1\textwidth]{images/sim1_FDR_all_models.png}
\caption{}
\label{fig:sim1_fdr}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/sim3_Accuracy.png}
\caption{}
\label{fig:sim2_acc}
\end{subfigure}
\caption{Dataset 1 \& 2: Computer simulation. a) In a clean simulation, all the considered methods in the first computer simulation correctly calibrated FDR when adjusting for changes in protein abundance. In comparison, the methods without accounting for the protein-level changes resulted in off-target, high false positive rates. b) In a noisy simulation, including limited feature observations and missing values, the advantage of the proposed approach was apparent. Looking at accuracy, the proposed method outperformed ANOVA and Limma in nearly every model.
}
\label{fig:computer_sim}
\end{figure}

### 2022-07-28 supplementary/ptm_sm.tex
addressed all comments in main

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim_new/spike_in_statistic_comparison}
\caption{Comparing the summary statistics between methods, the proposed method performed the strongest. In terms of accuracy and specificity the three methods were close, with Limma and Anova showing slightly higher values. Accuracy and specificity were dominated by the large number of true negatives (background peptides) compared to the true positives (spike-in peptides). In terms of recall, the proposed approach out performed the other two methods, showing that it correctly labeled the most spike-in peptides. \label{fig:spike_stat}}
\end{figure}

>>>>

%\begin{figure}[h!]
%\centering
%\includegraphics[width=.85\textwidth]{sim_new/spike_in_statistic_comparison}
%\caption{Comparing the summary statistics between methods, the proposed method performed the strongest. In terms of accuracy and specificity the three methods were close, with Limma and Anova showing slightly higher values. Accuracy and specificity were dominated by the large number of true negatives (background peptides) compared to the true positives (spike-in peptides). In terms of recall, the proposed approach out performed the other two methods, showing that it correctly labeled the most spike-in peptides. \label{fig:spike_stat}}
%\end{figure}
